An Extension of LTL with Rules\\ and its Application to Runtime Verification

### 2019-02-05 rule-ltl.tex
Updates from Overleaf

\doron{Did my best to make the letters more self explained. I think R is usually reserved in temporal logic for realtime. But if you insist, one thing that I cannot perform from here is changing the two figures where these letters occur.}

@@

\subsubsection{Extending LTL with Dynamic Quantification.}
Adding quantification over propositions allows writing a formula of the form
$\Exists q \, \varphi$, where
$\Exists q$ repesents {\em existential} quantification over a propositiona $q$ that can appear in $\varphi$.
Let the propositions $P$ be partitioned into two disjoint sets: 
{\em auxiliary propositions} $A = \{ a_1 , \; \ldots\; , a_n \}$
and {\em basic propositions} $B$. 
%A model for such a formula 
%has truth assignments only over the propositions in $B$, but the propositions $A$ can appear (only) within the scope of a quantifier.

To define the semantics, let $X \subseteq P$ 
%be a subset of 
%the propositional variables appearing in the %states of the sequence $\sigma = s_1 s_2 %\ldots$ 
and denote $\sigma |_X = s_1 \setminus X \, s_2 \setminus X \ldots$. (Note that $\sigma |_X$ denotes projecting {\em out} the variables in $X$.)

@@

We partition again, as in QLTL, the propositions $P$ into
{\em auxiliary propositions} $A = \{ a_1 , \; \ldots\; , a_n \}$
and {\em basic propositions} $B$. A model for an ELTL formula has truth assignments only over the propositions in $B$.
An ELTL property $\eta$ has the following form: 
\begin{equation} \label{ELTL}
\psi \mathit{\ where\ } a_j  := \varphi_j \, : \, 
 {j \in \{1, \ldots , n\}} \end{equation}
where each $a_i$ is a distinct auxiliary proposition from $A$,
$\psi$ is an LTL property and each $\varphi_i$ 
is a PLTL property where propositions from $B$ can
only occur within the scope of a $\ominus$ operator.
The semantics can be defined as follows. We refer to $\psi$
as the {\em statement} of $\eta$ and to 
$a_j  := \varphi_j \, : \, 
 {j \in \{1, \ldots , n\}}$ as the rules. In text, rules
 will be separated from each other by a comma (",").

@@

ELTL extends the set of propositional variables with new variables whose values at a state are
based on the prefix of the model. We say that
the extension of the states are {\em well founded},
as the added propositions are dependent and effectively summarize the past execution prefix up to and
including the current state.
This differs from 
the use of auxiliary variables in QLTL, where the auxiliary variables do not have to extend the execution
states in a fixed way throughout the interpretation of 
the property over a model. The constraint that auxiliary variables appearing in the formulas $\varphi_i$ 
must occur within the scope of a $\ominus$ operator
is required to guarantee that the values
of the auxiliary variables are not conflicting, as in
$a_1 = \neg a_2$ and $a_2 = a_1$, which makes a formula
that uses $a_1$ and $a_2$ uniformly $\false$.

\begin{lemma} \label{fourone}
Let $\varphi$ be an ELTL formula as in~(\ref{ELTL}),
over the auxiliary propositions $A$ and basic propositions $B$, 
and let $\sigma$ be a model with states over the $B$, such that
$\sigma \models \psi$.
Then there is a unique
model $\sigma'$ such that $\sigma' |_A = \sigma$.

@@

\klaus{(1) is this lemma saying the right thing? (2) is the proof really a proof of that lemma, (3) 
should $\varphi$ be $\psi$, (4) now $\psi$ is not the whole spec (it is missing the rules). What do we call the whole thing? It is not a formula like the 
$\varphi$s and the $\psi$s. Do we need a special greek symbol for it?}

\noindent {\bf Sketch of proof.} By induction on the state index. Each auxiliary variable
is defined, via a PLTL formula by the current state of the model over the values of the propositions $B$, and by the values of $A \cup B$ in previous states. A PLTL
property depends on a state only on the prefix
of the model that precedes and includes that state.
Restricting the auxiliary variables to appear only within the scope of a previous-time operator guarantees that
their values at the current state cannot be conflicting. \qed

@@

\klaus{How do we know that the ``other direction'' below is possible?}

For the other direction, one can translate any QLTL property into an $\omega$ automaton\footnote{The past operators of a QLTL property do not change this fact, since the translation can be done first to second order monadic logic, which can be translated into an $\omega$-automaton.}.
We assume that the property is given as a deterministic Muller automaton $M = (S, i , \Sigma, \Delta, {\cal F})$ (a B\"{u}chi automaton may not always be deterministic), where $S$ is the set of states, $i \in S$ is the initial state, $\Sigma$ is the set of inputs, each represented as a Boolean combination of the variables $B$, $\Delta : S \times \Sigma \mapsto S$ is the transition function, and ${\cal F} = \{ F_1 , \ldots , F_k \}$ is the accepting set, 
where an accepted input needs to pass infinitely often in exactly
one of the specified sets. To formulate an ELTL formula,
he auxiliary variables represent the states $S$ of $M$, i.e., $a_s$ represents the state $s \in S$.

\klaus{The following formulation above is not not clear: ``where an accepted input needs to pass infinitely often in exactly
one of the specified sets''.}

\noindent
Then we translate $M$ into an ELTL formula
of the form:

\begin{tabbing}
\ \ \ \ \ $\bigvee_{F \in {\cal F}} (\bigwedge_{s \in F} \Box\Diamond a_s \wedge \bigwedge_{s \not\in F} \Diamond \Box \neg a_s)$
where \\
\ \ \ \ \ \ \ \ \ \ $a_s := \vee _{\Delta ( s'
, \alpha ) = s } 
\alpha \wedge \ominus a_{s'} )
\wedge (s=i \rightarrow \neg \ominus \true)
\, :\,  s \in S
$
\end{tabbing}

\klaus{Above: should $s=i \rightarrow \neg \ominus \true$ instead be (note: two changes):
$\neg \ominus \true \rightarrow a_s=i$
}

\klaus{Above: can't judge whether the $\bigwedge_{s \not\in F} \Diamond \Box \neg a_s)$ is correct.}

\noindent
With $(s=i)$ is $\true$ when $s$ is $i$, and $\false$ otherwise.
Basically, the assignments to the auxiliary variables are used to simulate the change of states\footnote{It is possible to use only $\lceil \log ( | S | ) \rceil$
variables, at the expense of longer and more complicated formulas.}, and the main formula expresses the Muller acceptance condition. \qed

The use of the auxiliary variables $a_1 \ldots a_n$ in the
rules of an ELTL specification 
$\eta$ can be viewed as representing a
state machine, in particular when the only modality that is used in the rules part is $\ominus$.
In that case, we can consider an ELTL specification $\eta$ as consisting with two synchronous parts: an LTL property 
(the statement) $\psi$
and a finite automaton, whose states correspond to combinations
of truth values of the auxiliary variables, and can be referred to
within $\psi$. As a corollary of Theorem~\ref{theo1}, 
this restricted representation is also equivalent to QLTL.
This representation can be useful, for RV, as explored further in this paper, because of the possibility to
incrementally calculate the values of the auxiliary variables. 
Similarly, it can be
used for extending the expressiveness of LTL model checking.

\klaus{First sentence below sounds wrong: (1) PLTL already does not have future time operators, and (2) why mention infinite sequences?}

We can extend PLTL into EPLTL by disallowing the future temporal operators (except a prefixing $\Box$, when interpreted over infinite sequences). We obtain
a formalism that is equivalent to finite automata. In this case, 
we replace the first part of the construction in the proof of Theorem~\ref{theo1}
with the simpler formula $\vee_{s \in A} a_s$, where $A$ are the accepting states of the finite automaton.

\klaus{Last sentence above: perhaps say with words that we are always in an acceptance state - if that's what the point is.}

For representing safety properties that can be expressed in QPLTL, we can use a restricted version as shown in the following simple Lemma.
\begin{lemma}
\label{simplecase}
Any safety QPLTL property can be expressed using the following restricted
ELTL form:
\[  \Box p {\ where\ } a_j  = \varphi_j \, : \, 
 {j \in \{1, \ldots , n\}}  \] 
 with $p$ being one of the auxiliary variables $a_j$.
\end{lemma}
\noindent {\bf Proof.} The proof of the Lemma follows from the
axiomatization of QLTL in~\cite{MP}, e.g.,
$\varphi\ \since\ \psi \leftrightarrow \psi \vee ( \varphi \wedge (\varphi\ \since\ \psi))$. \qed

@@

While this Lemma makes the use of the temporal operators
besides $\ominus$ redundant, we believe that it is still
more convenient to use a combination of
temporal operators when possible, and
auxiliary variables definitions.
The relative expressive power between the propositional temporal logics discussed here
appear in Figure~\ref{prop}.

@@

Runtime verification of temporal specifications often
concentrates on the past portion of the logic.  Past-time  specifications have the important property that one can distinguish when they are violated. For an extended discussion of this issue of {\em monitorability}, see e.g.,~\cite{Ugly,FFM}.

The algorithm for PLTL, first presented in \cite{HR}, 
is based on the observation that the semantics of the 
past time formulas $\ominus \varphi$ and $(\varphi\, \since \,\psi)$ in the current step $i$ is defined in terms of the semantics
in the previous step $i - 1$ of a subformula.
To demonstrate this, we repeat here the definition of the previous-time
operator $\ominus$, and rewrite the definition
of $\since$ in an equivalent form that is
more directly amenable for runtime verification.

\begin{itemize}
\item $( \sigma , i) \models \ominus \varphi$ iff $i > 1$ and $(\sigma, i-1) \models \varphi$.
\item $(\sigma , i) \models (\varphi \, \since \, \psi)$ iff $(\sigma , i) \models \psi$ or the following hold: $i>1$,
$( \sigma , i)  \models \varphi$ and 
$(\sigma , i-1 ) \models ( \varphi \, \since \, \psi )$.
\end{itemize}

\noindent
Runtime verification exploits the fact that
the semantic definition is recursive in both the structure of the property, where subformulas
are evaluated based on smaller subformulas, and the
evaluation of subformulas in the previous state.
The algorithm, shown below, operates on two vectors (arrays) of values indexed by subformulas:  $\old$ that summarizes the truth values of the
subformulas just {\em before} the new state, and $\current$ for the {\em current} state.

\begin{enumerate}
\item Initially, for each subformula $\varphi$,
%$B ( \varphi , \current ) = \bfalse $.
$\current ( \varphi ) := \false$.
\item Observe a new event (as a set of ground predicates) $s$ as input. 
\item Let $\old := \current$.
\item Make the following updates for each subformula. If $\varphi$ is
      a subformula of $\psi$ then $\current ( \varphi )$ is updated before 
      $\current ( \psi )$.

@@

\noindent
Runtime verification of QPLTL can be done by translating a
formula into a deterministic finite automaton. Then the sequence of states to be checked forms the input for this automaton. 
However, QPLTL can be extremely compact, in particular when using alternating existential and universal 
quantifiers, which can result in a huge (non-elementary) explosion of
the automaton representing the same
specification~\cite{Thomas}.

The well foundedness of the formulas defining the new auxiliary variables in EPLTL makes the extension of the above
runtime algorithm simple.
In order to extend the algorithm to perform runtime verification of an EPLTL formula $\psi$, as defined in~(\ref{ELTL}),  we need to  add to the above
algorithm memory for $\current ( a_j )$ and $\old ( a_j )$,
and perform the following between steps 3 and 4 for each $a_j$: \\

\ \ \ \ \ $\current ( a_j ) := \current ( \varphi_j )$ \\

\subsubsection{Evaluation Order}

The order of calculating $\current$ is subtle here, because
auxiliary variables can appear recursively on the right-hand 
side of their definition. For PLTL, one can start calculating the subformulas
bottom up or top down. But for EPLTL, this would not work. To see this, consider, for example, the formula~(\ref{form3}). It contains
the definition $q := \ominus \neg q$. Now, we cannot calculate this
bottom up, since the left-hand $q$ is not calculated yet, and we need its value to compute the right-hand side $q$ leaf node.
If we calculate it top down, we will encounter the same problem.
However, notice that the top down calculation does not need the
value of $q$ to calculate $\ominus \neg q$; this value should already be available as $\old ( \neg q )$.

\klaus{I don't see that we need to talk about the top-down evaluation.}

@@

\begin{enumerate}
\item Calculate 
$\current ( \delta )$ for each subformula $\delta$
that appears in $\varphi_j$ of rule
$a_j := \varphi_j$ but not within the scope of
a $\ominus$ operator. (This is not
a problem since for $\delta = \ominus \gamma$,
$\current ( \delta )$ is set to $\old ( \gamma )$.)
\item Set 
$\current ( a_j )$ to $\current ( \varphi_j )$ for each $j$.
\item Calculate $\current ( \delta )$ for each subformula $\delta$
that appears in $\varphi_j$ of a rule
$a_j = \varphi_j$ {\em  within} the scope of
a $\ominus$ operator.
\item Calculate $\current ( \delta )$ for each subformula $\delta$
that appears in the statement $\psi$ - using $\current ( a_j )$.
\end{enumerate}

@@

Assume a finite set of infinite domains\footnote{Finite domains are handled with some minor changes, see~\cite{HPU}.}
$D_1, D_2, \ldots$,
e.g., integers or strings. 
Let $V$ be a finite set of {\em variables}, 
with typical instances $x$, $y$, $z$.
%We denote by $x: D$ the fact that the variable $x$ has the domain $D$.
An {\em assignment} over a set of variables $V$
maps each variable $x \in V$ to a value from
its associated domain $\domain ( x )$, where multiple variables (or all of them)
can be related to the same domain. For example
$[ x \rightarrow 5 , y \rightarrow \text{``abc''} ]$ assigns
the values $5$ to $x$ and the value ``abc'' to $y$.
Let $T$ be a set of {\em relations} 
with typical instances $R$.
Each relation is associated with some position-dependent domains.

\klaus{Above: let T be ... Is T used later? The relations are used in the next paragraph, why not introduce them there? Also, relations are not defined really.}

\klaus{I don't quite understand the motivation below for this way of modeling events: relations with a time stamp, and the whole story of the time stamp not being mentioned etc. Why not just the usual models: traces? Too late to change now for CAV though.}

We define models for FLTL based on {\em temporal} relations~\cite{Chomicki}. That is,
their last parameter
is a natural number. For a relation $R$, $R [ i ]$ is the relation obtained from $R$ by
restricting it to the value $i$ in the last parameter. For simplicity, we will describe henceforth
the logic with relations $R$ that have exactly two parameters, 
hence $R [ i ]$ will have just one parameter whose domain will be denoted as $d ( R )$; this domain is fixed over all the values of the second parameter $i$. The definition of the logic over other numbers of parameters is quite straightforward, and our implementation of runtime verification fully supports it.

\klaus{Above, what about instead of writing: ``this domain is fixed over all the values of the second parameter $i$'' to write:
``this domain is independent of the values of the second parameter $i$'' - if that is what you mean to say.}

\klaus{Above: we need to say explicitly what $i$ intuitively is meant to represent: positions, time, ....}

@@

Additional operators are defined as in the propositional logic. In addition we define
$\forall x \; \varphi = \neg \exists x \neg \varphi$.
Restricting the modal operators to the past ones
($\since$, $\ominus$ and the ones obtained from them) 
forms the logic PFLTL.
The quantification over values of variables here is `static' in the sense that values quantified over are independent of the state in the execution. We
denote static quantification with $\exists$ and $\forall$, while
dynamic quantification is expressed with $\Exists$ and $\Forall$
(see e.g. Section \ref{sec:extending-prop-ltl}).

@@

A model is a set of
 relations ${\cal R} = \{ R_1 \, \ldots , R_m \}$, where each of these relations has a last parameter $i$ that is a natural number. 
% 
It is standard, and more intuitive (in particular for runtime verification) to cast models of temporal logic in terms of sequences states rather than temporal relations. To do that,
 we denote by $R_k [ i ]$ the restriction of
 the relation $R_k$ to the value $i$ of the last parameter. A model ${\cal R}$ is then a sequence ${\cal R} [ 1 ]\, {\cal R} [ 2 ] \ldots$, where
 each {\em state} ${\cal R} [i] = \{ R_1 [i], \ldots , R_m [ i ] \}$  (a set of relations).
 The denotation $p_k$ will correspond to the relation $R_k$.

\klaus{Above, did you not already say the following before:
``we denote by $R_k [ i ]$ the restriction of
 the relation $R_k$ to the value $i$ of the last parameter.''.}

@@

We now extend FLTL into EFLTL in a way that is motivated by the  propositional extension from LTL (PLTL) to ELTL (EPLTL). We thus allow the following formula:

\begin{equation}
\label{EFLTL}
\psi \mathit{\ where\ } r_j  ( x_j ) := 
\varphi_j (x_j) : j \in \{ 1 , \ldots , n \} 
\end{equation}

\begin{enumerate}
\item $\psi$, the {\em statement}, is a FLTL formula with
no free variables, 
\item $\varphi_j$ are PFLTL formulas with a single
free variable $x_j$,
\item $r_j$ is an auxiliary temporal relation with $2$ parameters; the first parameter is of the same 
type as $x_j$ and the second one is, as usual,
a natural number that is omitted in the temporal formulas. 
An auxiliary relation $r_j$ can appear within $\psi$. It can also appear in $\varphi_k$ of a {\em rule} $r_k := \varphi_k$, but only within the
scope\footnote{This restriction has the same motivation as in
the propositional version ELTL (and EPLTL): to avoid making the
formula $\false$ just by means of the definition of the auxiliary relations.}
of a previous-time operator $\ominus$. 
\end{enumerate}

>>>>

%\doron{Did my best to make the letters more self explained. I think R is usually reserved in temporal logic for realtime. But if you insist, one thing that I cannot perform from here is changing the two figures where these letters occur.}

@@

\subsubsection{Extending LTL with Dynamic Quantification.}
Adding quantification over propositions allows writing a formula of the form
$\Exists q \, \varphi$, where
$\Exists q$ repesents {\em existential} quantification over a propositiona $q$ that can appear in $\varphi$.

%A model for such a formula 
%has truth assignments only over the propositions in $B$, but the propositions $A$ can appear (only) within the scope of a quantifier.
To define the semantics, let $X \subseteq P$ 
%be a subset of 
%the propositional variables appearing in the %states of the sequence $\sigma = s_1 s_2 %\ldots$
and denote $\sigma |_X = s_1 \setminus X \, s_2 \setminus X \ldots$. (Note that $\sigma |_X$ denotes projecting {\em out} the variables in $X$.)

@@

We partition the propositions $P$ into
{\em auxiliary propositions} $A = \{ a_1 , \; \ldots\; , a_n \}$
and {\em basic propositions} $B$. A model for an ELTL formula has truth assignments only over the propositions in $B$.
An ELTL property $\eta$ has the following form: 
\begin{equation} \label{ELTL}
\psi \mathit{\ where\ } a_j  := \varphi_j \, : \, 
 {j \in \{1, \ldots , n\}} \end{equation}
where each $a_i$ is a distinct auxiliary proposition from $A$,
$\psi$ is an LTL property and each $\varphi_i$ 
is a PLTL property where propositions from $B$ can
only occur within the scope of a $\ominus$ operator.
The semantics can be defined as follows. We refer to $\psi$
as the {\em statement} of $\eta$ and to 
$a_j  := \varphi_j \, : \, 
 {j \in \{1, \ldots , n\}}$ as the {\em rules}. In text, rules
 will be separated from each other by a comma (",").

@@

ELTL extends the set of propositional variables with new variables whose values at a state are
based on the prefix of the model. The added propositions are dependent and effectively summarize the past execution prefix up to and
including the current state.
This differs from 
the use of auxiliary variables in QLTL, where the auxiliary variables do not have to extend the execution
states in a fixed way throughout the interpretation of 
the property over a model. The constraint that auxiliary variables appearing in the formulas $\varphi_i$ 
must occur within the scope of a $\ominus$ operator
is required to help preventing that the values
of the auxiliary variables are not conflicting, as in
$a_1 := \neg a_2$ and $a_2 := a_1$, which makes a formula
that uses $a_1$ and $a_2$ uniformly $\false$. Still, one can write conflicting rules, e.g., by simlpy 
adding the rule $a: = \false$.

\begin{lemma} \label{fourone}
Let $\varphi$ be an ELTL formula
over the auxiliary propositions $A = \{ a_1 , \ldots , a_n \}$ and basic propositions $B$, with
a set of rules $a_j  := \varphi_j \, : \, 
 {j \in \{1, \ldots , n\}}$. 
Let $\sigma$ be a model with states over the $B$. 
Then there is at most one
model $\sigma'$ such that $\sigma' |_A = \sigma$ and $\sigma' \models \Box \bigwedge_{1 \leq j \leq n} ( a_j \leftrightarrow \varphi_j) $

@@

%\klaus{(1) is this lemma saying the right thing? (2) is the proof really a proof of that lemma, (3) 
%$\varphi$s and the $\psi$s. Do we need a special greek symbol for it?}

\noindent {\bf Sketch of proof.} By induction on the state index. Each auxiliary variable
is defined, via a PLTL formula by the current state of the model over the values of the propositions $B$, and by the values of $A \cup B$ in previous states. A PLTL
property depends in a state only on the prefix
of the model that precedes and includes that state. \qed

@@

%\klaus{How do we know that the ``other direction'' below is possible?}

For the other direction, one can translate the QLTL property first into
a second order monadic logic formula, then to a deterministic Muller automata and then constructs an ELTL formula that holds for the accepting executions of this automaton. The details of the
constructions are given in the appendix. \qed

%EPLTL, where the Muller automaton can be
%replaced with a B\"{u}chi automaton,
%where all of its states, except one, sink
%state, are accepting. \qed

\iffalse
The use of the auxiliary variables $a_1 \ldots a_n$ and the
rules of an ELTL specification 
can be viewed as representing a
state machine, in particular when the only modality that is used in the rules part is $\ominus$.
In that case, we can consider an ELTL specification $\eta$ as consisting of two synchronous parts: an LTL property 
(the statement) 
and a finite automaton, whose states correspond to combinations
of truth values of the auxiliary variables, and can be referred to
within $\psi$. As a corollary of Theorem~\ref{theo1}, 
this restricted representation is also equivalent to QLTL.
This representation can be useful, for RV, as explored further in this paper, because of the possibility to
incrementally calculate the values of the auxiliary variables. 
Similarly, it can be
used for extending the expressiveness of LTL model checking.

%\klaus{First sentence below sounds wrong: (1) PLTL already does not have future time operators, and (2) why mention infinite sequences?}
\fi

We extend PLTL into EPLTL by disallowing the future temporal operators (except an implied prefixing $\Box$). This results in
a formalism that is equivalent to finite automata, where all the states except one are accepting and where the nonaccepting state is a sink. 
We can use related, but simpler construction than in Theorem~\ref{theo1} to prove the following:

\iffalse
In this case, 
we replace the first part of the construction in the proof of Theorem~\ref{theo1}
with the simpler formula $\vee_{s \in A} a_s$, where all the states but one are
accepting.$A$ are the accepting states of the finite automaton.

\klaus{Last sentence above: perhaps say with words that we are always in an acceptance state - if that's what the point is.}
\fi

%For representing safety properties that can be expressed in QPLTL, we can use a restricted version as shown in the following simple Lemma.

\begin{lemma}
\label{simplecase}
The expressive power of 
ELTL is the same as QLTL.
%Any safety QPLTL property can be expressed using the following restricted
%ELTL form:
%\[  \Box p {\ where\ } a_j  = \varphi_j \, : \, 
% {j \in \{1, \ldots , n\}}  \] 
% with $p$ being one of the auxiliary variables %$a_j$.
\end{lemma}
In fact, it is easy to show that EPLTL can be restricted to the following form, while retaining its expressive power.
\[  \Box p {\ where\ } a_j  = \varphi_j \, : \, 
{j \in \{1, \ldots , n\}}  \] 
with $p$ being one of the auxiliary variables $a_j$. The relative expressive power between the propositional temporal logics discussed in this section is summarized in Figure~\ref{prop}.

\iffalse
\noindent {\bf Proof.} The proof of the Lemma follows from the
axiomatization of QLTL in~\cite{MP}, e.g.,
$\varphi\ \since\ \psi \leftrightarrow \psi \vee ( \varphi \wedge (\varphi\ \since\ \psi))$. \qed

@@

While this Lemma makes the use of the temporal operators
besides $\ominus$ in EPLTL redundant, it is still
convenient to use a combination of
temporal operators when possible, and
auxiliary variables definitions.
\fi

@@

Runtime verification of temporal specification often
concentrates on the past portion of the logic.  Past-time  specifications have the important property that one can distinguish when they are violated. For an extended discussion of this issue of {\em monitorability}, see e.g.,~\cite{Ugly,FFM}.

The algorithm for PLTL, first presented in \cite{HR}, 
is based on the observation that the semantics of the 
past time formulas $\ominus \varphi$ and $(\varphi\, \since \,\psi)$ in the current step $i$ is defined in terms of the semantics
in the previous step $i - 1$ of a subformula.
To demonstrate this, we repeat here the definition of the previous-time
operator $\ominus$, and rewrite the definition
of $\since$ in an equivalent form that is
more directly applicable for runtime verification.
\begin{itemize}
\item $( \sigma , i) \models \ominus \varphi$ iff $i > 1$ and $(\sigma, i-1) \models \varphi$.
\item $(\sigma , i) \models (\varphi \, \since \, \psi)$ iff $(\sigma , i) \models \psi$ or the following hold: $i>1$,
$( \sigma , i)  \models \varphi$ and 
$(\sigma , i-1 ) \models ( \varphi \, \since \, \psi )$.
\end{itemize}
Runtime verification exploits the fact that
the semantic definition is recursive in both the structure of the property, where subformulas
are evaluated based on smaller subformulas, and the
evaluation of subformulas in the previous state.
The algorithm, shown below, operates on two vectors (arrays) of values indexed by subformulas:  $\old$ that summarizes the truth values of the
subformulas just {\em before} the new state, and $\current$ for the {\em current} state. The order of calculating $\current$ for subformulas is bottom up, according to the syntax tree.
\begin{enumerate}
\item Initially, for each subformula $\varphi$,
%$B ( \varphi , \current ) = \bfalse $.
$\current ( \varphi ) := \false$.
\item Observe a new event (as a set of ground predicates) $s$ as input. 
\item Let $\old := \current$.
\item Make the following updates for each subformula. If $\varphi$ is
      a subformula of $\psi$ then $\current ( \varphi )$ is updated before 
      $\current ( \psi )$.

@@

\subsubsection*{Runtime verification for QPLTL.}  This can be performed by translating a
formula into a deterministic finite automaton. Then the sequence of states to be checked forms the input for this automaton. 
However, QPLTL can be extremely compact, in particular when using alternating existential and universal 
quantifiers, which can result in a huge (non-elementary) explosion of
the automaton representing the same
specification~\cite{Thomas}.

\subsubsection*{Runtime verification for EPLTL.} For EPLTL,  we need to  add to the above
algorithm calculations for the rules of
the form $a_j := \varphi_j$ (calculations
for the statement part $\psi$ remain the same).
$\current ( a_j )$ and $\old ( a_j )$,
and perform the following between steps~3 and~4 for each $a_j$: $\current ( a_j ) := \current ( \varphi_j )$.

For PLTL, we can calculate  $\current$, for subformulas,
bottom up. But
because
auxiliary variables can appear recursively in EPLTL rules, i.e., both on the left and the right sides,
this would not work correctly. To see this, consider, for example, the formula~(\ref{form3}). It contains
the definition $q := \ominus \neg q$. Now, we cannot calculate this
bottom up, since the left-hand $q$ is not calculated yet, and we need its value to compute the right-hand side $q$ leaf node.
If we calculate it top down, we will encounter the same problem.
However, notice that the top down calculation does not need the
value of $q$ to calculate $\ominus \neg q$; this value should already be available as $\old ( \neg q )$.

%\klaus{I don't see that we need to talk about the top-down evaluation.}

@@

\begin{enumerate}
\item Calculate 
$\current ( \delta )$ for each subformula $\delta$
that appears in $\varphi_j$ of a rule
$a_j := \varphi_j$, but {\em not} within the scope of
a $\ominus$ operator. (This is not
a problem, since
$\current ( \ominus \gamma )$ is set to $\old ( \gamma )$.)
\item Set 
$\current ( a_j )$ to $\current ( \varphi_j )$ for each $j$.
\item Calculate $\current ( \delta )$ for each subformula $\delta$
that appears in $\varphi_j$ of a rule
$a_j = \varphi_j$ {\em  within} the scope of
a $\ominus$ operator.
\item Calculate $\current ( \delta )$ for each subformula $\delta$
that appears in the statement $\psi$, using the calculated $\current ( a_j )$.
\end{enumerate}

@@

Assume a finite set of infinite domains\footnote{Finite domains are handled with some minor changes, see~\cite{HPU}.}
$D_1, D_2, \ldots$,
e.g., integers or strings. 
Let $V$ be a finite set of {\em variables}, 
with typical instances $x$, $y$, $z$.
%We denote by $x: D$ the fact that the variable $x$ has the domain $D$.
An {\em assignment} over a set of variables $V$
maps each variable $x \in V$ to a value from
its associated domain $\domain ( x )$, where multiple variables (or all of them)
can be related to the same domain. For example
$[ x \rightarrow 5 , y \rightarrow \text{``abc''} ]$ assigns
the values $5$ to $x$ and the value ``abc'' to $y$.
%Let ${\cal R}$ be a set of {\em relations} 
%with typical instances $R$.
%Each relation is associated with some position-dependent domains.

%\klaus{Above: let T be ... Is T used later? The relations are used in the next paragraph, why not introduce them there? Also, relations are not defined really.}
%\klaus{I don't quite understand the motivation below for this way of modeling events: relations with a time stamp, and the whole story of the time stamp not being mentioned etc. Why not just the usual models: traces? Too late to change now for CAV though.}

We define models for FLTL based on {\em temporal} relations~\cite{Chomicki}. That is relations with
last parameter
that is a natural number represented a time instance. So a tuple of
a relation $R$ can be $(``a'', 5, ``cbb'', 3)$, where
the $3$ is the time parameter. The last parameter $i$
does not correspond to modeling physical time. This
is rather used to allow the relations to have
different tuples in different instances of $i$,
corresponding to states, in the propositional temporal
logics, so that we can define the temporal operators
such as $\bigcirc$ or $\since$.

For a relation $R$, $R [ i ]$ is the relation obtained from $R$ by
restricting it to the value $i$ in the last parameter, and removing that last $i$ from the tuples. For simplicity, we will describe henceforth
the logic with relations $R$ that have exactly two parameters, 
hence $R [ i ]$ will have just one parameter whose domain will be denoted as $d ( R )$. The definition of the logic over other numbers of parameters is quite straightforward. e.g., when $R$ has one parameter,
$R [ i ]$ is a Boolean value. Our implementation of runtime verification described later fully supports relations with zero or more parameter.

%\klaus{Above, what about instead of writing: ``this domain is fixed over all the values of the second parameter $i$'' to write:
%``this domain is independent of the values of the second %parameter $i$'' - if that is what you mean to say.}

%\klaus{Above: we need to say explicitly what $i$ intuitively is meant to represent: positions, time, ....}

@@

Additional operators are defined as in the propositional logic. We שךדם define
$\forall x \; \varphi = \neg \exists x \neg \varphi$.
Restricting the modal operators to the past operators
($\since$, $\ominus$ and the ones obtained from them) 
forms the logic PFLTL.
The quantification over values of variables here is `static' in the sense that they are independent of the state in the execution. We
denote static quantification with $\exists$ and
$\forall$.
Dynamic quantification, as defined for LTL in Section~\ref{sec:extending-prop-ltl},
and will extend FLTL below, is expressed with $\Exists$ and $\Forall$.

@@

A model is a set of temporal
 relations ${\cal R} = \{ R_1 \, \ldots , R_m \}$.
 Since the standard definition of temporal logic is
 over a sequence (``the execution''). let
 ${\cal R} [ i ] = \{ R_1 [ i ] \, \ldots , R_m [ i ] \}$. A model ${\cal R}$ can thus be
seen as the sequence ${\cal R} [ 1 ]\, {\cal R} [ 2 ] \ldots$.

%\klaus{Above, did you not already say the following before:
%``we denote by $R_k [ i ]$ the restriction of
% the relation $R_k$ to the value $i$ of the last parameter.''.}

@@

We now extend FLTL into EFLTL in a way that is motivated by the  propositional extension from LTL (PLTL) to ELTL (EPLTL). We thus allow the following formula:
\begin{equation}
\label{EFLTL}
\psi \mathit{\ where\ } r_j  ( x_j ) := 
\varphi_j (x_j) : j \in \{ 1 , \ldots , n \} 
\end{equation}
\begin{enumerate}
\item $\psi$, the {\em statement}, is a FLTL formula with
no free variables, 
\item $\varphi_j$ are PFLTL formulas with a single
free variable $x_j$,
\item $r_j$ is an auxiliary temporal relation with $2$ parameters; the first parameter is of the same 
type as $x_j$ and the second one is, as usual,
a natural number that is omitted in the temporal formulas. 
An auxiliary relation $r_j$ can appear within $\psi$. It can also appear in $\varphi_k$ of a {\em rule} $r_k := \varphi_k$, but only within the
scope\footnote{This restriction has the same motivation as in
the propositional version ELTL (and EPLTL): to avoid making the
formula $\false$ just by means of the definition of the auxiliary relations.}
of a previous-time operator $\ominus$. 
\end{enumerate}
